The Implicit Calculus: A New Foundation for Generic Programming

### 2017-01-24 src/Translation.tex
updating Sections 3 - 4: work in progress

\begin{comment}

\figtwocol{f:trans}{Type-directed Translation to System F}{
\small
\bda{llrl} 
\text{Type Environments} & \tenv & ::= & \epsilon \mid \tenv,
\relation{x}{\rulet} \\
\text{Singleton Environments} & \Theta & ::= & \rulet \leadsto E\\
\text{Translation Environments} & \env & ::= & \epsilon \mid \env, \rulet \leadsto x \\
\eda \\
\bda{lc}

& \multicolumn{1}{c}{
  \myruleform{\tenv \mid \denv \turns \relation{e}{\rulet} \leadsto E}} \\
\\
\TrInt &
{ \tenv \mid \denv \turns \relation{n}{\tyint} \leadsto n } 
\\ \\

\TrVar &
\myirule
{ (\relation{x}{\rulet}) \in \tenv}
{ \tenv \mid \denv \turns \relation{x}{\rulet} \leadsto x
} 
\\ \\

\TrAbs &
\myirule
{ \tenv;\relation{x}{\rulet_1}\mid \denv \turns \relation{e}{\rulet_2} \leadsto E
}
{ \tenv\mid\denv \turns \relation{\lambda \relation{x}{\rulet_1}.e}{\rulet_1 \arrow \rulet_2}
  \leadsto \lambda \relation{x}{|\rulet_1|}.E } 
\\ \\

\TrApp &
\myirule
{ \tenv\mid \denv \turns \relation{e_1}{\rulet_1 \arrow \rulet_2} \leadsto E_1 \\
  \tenv\mid \denv \turns \relation{e_2}{\rulet_1} \leadsto E_2
}
{ \tenv\mid\denv \turns \relation{e_1\,e_2}{\rulet_2} \leadsto E_1\,E_2} 
\\ \\

\TrTAbs&
  \myirule { \tenv,\alpha \mid \env \turns \relation{e}{\rho} \leadsto
    E_1 }
           { \tenv\mid\env \turns \relation{\Lambda \alpha.e}{\forall
               \alpha.\rho} \leadsto \Lambda \alpha.E_1 } \quad\quad\quad
\\ \\
\TrTApp&
  \myirule { \tenv\mid \env \turns \relation{e}{\forall \alpha.\rho_2} \leadsto
    E}
           { \tenv\mid\env \turns \relation{e\,\rho_1}{\rho_2 [\rho_1 /\alpha]} \leadsto
    E~|\rho_1|} 
\\ \\
\TrIAbs&
  \myirule { \tenv \mid \env, \rho_1 \leadsto x \turns \relation{e}{\rho_2}  \leadsto
    E}
           { \tenv\mid\env \turns \relation{\ilambda \rho_1.e}{\rho_1 \iarrow \rho_2}  \leadsto
    \lambda \relation{x}{|\rulet_1|}. E}
\\ \\
\TrIApp&
  \myirule { \tenv \mid \env \turns \relation{e_1}{\rho_2 \iarrow
      \rho_1 \leadsto E_1} \\
             \tenv \mid \env \turns \relation{e_2}{\rho_2} \leadsto E_2}
           { \tenv\mid\env \turns \relation{e_1 \with e_2}{\rho_1} \leadsto
    E_1~E_2}
\\ \\
\TrQuery &
\myirule
{ \denv \vturns \rulet \leadsto E}
{ \tenv\mid\denv \turns \relation{?\rulet}{\rulet} \leadsto E
}

%%\TrRule &
%%\myirule
%%{ \rulet = \rulesch{\alpha}{\ruleset}{\type} \quad
%%  \vec{\alpha} \cap \mathit{ftv}(\tenv,\env) = \emptyset \\
%%  \tenv\mid\denv; \overline{\relation{\rulet}{x}} \turns 
%%  \relation{e}{\type} \leadsto E \quad\quad
%%  \bar{x}\text{~fresh} 
%%}
%%{ \tenv\mid\denv \turns 
%%  \relation{\ruleabs{\rulet}{e}}{\rulet} \leadsto
%%  \Abs{\vec{\alpha}}{\abs{({\relation{\vec{x}}{|\vec{\rulet}|}})}{E}}
%%} 
%%\\ \\

%%\TrInst &
%%\myirule
%%{ \tenv\mid\denv \turns \relation{e}{\rulesch{\alpha}{\rulesetvar}{\type}} \leadsto E
%%}
%%{ \tenv\mid\denv \turns 
%%  \relation
%%  {e[\vec{\type}]}
%%  {\subst{\vec{\alpha}}{\vec{\type}}(\rulesetvar \To \type)} 
%%  \leadsto E\;|\vec{\type}|
%%  
%%} 
%%\\ \\

%%\TrRApp &
%%\myirule
%%{ \tenv\mid\denv \turns \relation{e}{\ruleset \To \type} \leadsto E \\
%%  \tenv\mid\denv \turns \relation{e_i}{\rulet_i} \leadsto E_i \quad 
%%  (\forall \relation{e_i}{\rulet_i} \in \rulesetexp)
%%}
%%{ \tenv\mid\denv \turns \relation{(\ruleapp{e}{\rulesetexp})}{\type}
%%  \leadsto E~\vec{E}}
%%\\ \\

%%\multicolumn{2}{l}{\myruleform{\denv \vturns \rho \leadsto E}} \\ \\

%%\TrRes &
%%\myirule
%%{ 
%%  \denv(\type) = 
%%  \relation{\rulesetvar' \Rightarrow \type}{E} \quad\quad
%%  \bar{x}\text{~fresh} \\

%%  \forall \rulet_i \in \rulesetvar': 
%%  \left\{
%%    \begin{array}{ll}
%%      \denv \vturns \rulet_i \leadsto E_i &
%%      , \rulet_i \not \in \rulesetvar  \\
%%      E_i = x_i &
%%      , \rulet_i \in \rulesetvar
%%    \end{array}
%%  \right.
%%}
%%{ \denv \vturns \rulesch{\alpha}{\rulesetvar}{\type} \leadsto
%%  \Abs
%%  {\vec{\alpha}}
%%  {\abs
%%    {(\relation{\vec{x}}{|\vec{\rulet}|})}
%%    {(E\,\vec{E})}}
%%}

% \multicolumn{2}{c}{
%   \coherent(\denv, \type) \defeq 
%   \forall \theta. \theta \denv(\type) = (\theta \denv)(\theta \type) 
% }

%%\eda
%%\bda{llc}
%%\myruleform{\lookup{\env}{\type} = \rulet : E} &  
%%\myirule
%%{
%%  \lookup{\overline{\relation{\rulet}{x}}}{\type} = \rulet : E
%%}
%%{
%%  \lookup{(\env;\overline{\relation{\rulet}{x}})}{\type} = \rulet : E
%%} \\ \\
%% & 
%%\myirule
%%{
%%  \lookup{\overline{\relation{\rulet}{x}}}{\type} = \bot \quad 
%%  \lookup{\env}{\type} = \rulet
%%}
%%{
%%  \lookup{(\env;\overline{\relation{\rulet}{x}})}{\type} = \rulet
%%} \\ \\

%%\myruleform{\lookup{\overline{\relation{\rulet}{x}}}{\type} = \rulet : E} &
%% \myirule
%%{
%%  (\relation{\rulet}{x}) \in \overline{\relation{\rulet}{x}} \quad\quad \rulet = \ruleschr{\vec{\alpha}'}{\rulesetvar'}{\type'} \\
%%  \theta\type' = \type  \quad\quad \theta = [\vec{\alpha}' \mapsto \vec{\type} ]
%%}
%%{
%%  \lookup{\overline{\relation{\rulet}{x}}}{\type} = \theta\bar{\rulet}' \Rightarrow \type : x\,|\vec{\type}|
%%}

\eda \\
\begin{eqnarray*}
|\alpha| & = & \alpha \\
|\tyint| & = & \tyint \\
|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| \\
|\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
|\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow
|\rulet_2| %\\
%%|\type|_\rulet & = & |\type|_\type \\ 
%%|\forall \vec{\alpha}.\myset{\rulet_1, \cdots, \rulet_n} \To \type| & = & 
%%\forall \vec{\alpha}.|\rulet_1| \to \cdots \to |\rulet_n| \to |\type| \\
%|\tenv| & = & \myset{ (\relation{x}{|\rulet|})  ~|~ (\relation{x}{\rulet}) \in \tenv } \\
%|\denv| & = & 
%\myset{ (\relation{x}{|\rulet|}) ~|~ 
%  (\relation{\rulet}{x}) \in \denv }
\end{eqnarray*}
}

\figtwocol{fig:resolution2}{Deterministic Resolution Translation}{
\begin{center}
\framebox{$
\ba{c}
\myruleform{\env \vturns \rho \leadsto E}
\quad\quad\quad
\TrRTAbs \quad
  \myirule{\env, \rho_1 \leadsto x \vturns \rho_2 \leadsto E \quad\quad x~\mathit{fresh}}
          {\env \vturns \rho_1 \iarrow \rho_2 \leadsto
            \lambda\relation{x}{|\rho_1|}.E} 
\\ \\
\TrRIAbs \quad
  \myirule{\env \vturns \rho \leadsto E}
          {\env \vturns \forall \alpha. \rho \leadsto \Lambda\alpha.E} 
\quad\quad\quad
\TrRSimpl \quad
 \myirule{\env\langle\type\rangle = \Theta \quad\quad \env; \Theta
   \turns_\downarrow \type \leadsto E}
          {\env \vturns \type \leadsto E} 
\\ \\ \\
\myruleform{\env; \Theta \turns_\downarrow \type \leadsto E}
\quad\quad\quad
\TrIIAbs \quad
  \myirule{\env \vturns \rho_1 \leadsto E_2 \quad\quad \env; \rho_2
    \leadsto E_1~E_2
    \turns_\downarrow \type \leadsto E_3}
          {\env; \rho_1 \iarrow \rho_2 \leadsto E_1 \turns_\downarrow \type
            \leadsto E_3}  
\\ \\
\TrIDone \quad
  \myirule{}
          {\env; \type \leadsto E \turns_\downarrow \type \leadsto E}  
\quad\quad\quad
\TrITAbs \quad
  \myirule{\env; \rho[\rulet'/\alpha] \leadsto E_1~|\rulet'|\turns_\downarrow \type \leadsto
  E_2}
          {\env; \forall \alpha.\rho \leadsto E_1 \turns_\downarrow
            \type \leadsto E_2}  
\\ \\ \\
\myruleform{\env\langle\type\rangle = \Theta}
\quad\quad
\TrLHere \quad
  \myirule{\rulet \lhd \type}
          {(\env,\rulet \leadsto E)\langle\type\rangle = \rulet \leadsto E}  
\quad\quad
\TrLThere \quad
  \myirule{\rulet \mathop{\not\!\!\lhd} \type \quad\quad \env\langle\type\rangle = \Theta}
          {(\env,\rulet \leadsto E)\langle\type\rangle = \Theta}  
%%\\ \\
%%\myruleform{\rho\lhd\type} \\ \\
%%  \myirule{}
%%          {\type \leadsto E \lhd \type}
%%\quad\quad\quad
%%  \myirule{\rho[\rulet'/\alpha] \lhd \type}
%%          {\forall \alpha.\rho \lhd \type}
%%\quad\quad\quad
%%  \myirule{\rho' \lhd \type}
%%          {\rho'' \iarrow \rho' \lhd \type}
\ea
$
}
\end{center}
}

\end{comment}

@@

%-------------------------------------------------------------------------------
\subsection{Type-Directed Translation}
Figure~\ref{fig:type} presents the translation rules that convert $\ourlang$
expressions into ones of System F. 
The gray parts of the figure essentially extend the type system with the necessary
information for the translation.
%%, but for readability we have omitted the
%%earlier gray-shaded conditions.

@@

With respect to the type system the type environments $\Gamma$ remain
the same. However, implicit environments $\env$ need to be extended with
evidence information for the translation, thus becoming
\emph{translation} environments:

{\small
  \[ \begin{array}{llrl}
\text{Translation Environments} & \env & ::= & \epsilon \mid \env, \rulet~\gbox{\leadsto E} \\
  \end{array} \]}

The main translation judgment, which adapts the typing judgment, is 
\begin{center}
  $\tenv \mid \denv \turns \relation{e}{\rho}~\gbox{\leadsto E}$
\end{center}
This judgment states that the translation of $\ourlang$ expression $e$ with
type $\rho$ is System~F expression $E$, with respect to type environment
$\tenv$ and translation
environment $\denv$. The translation environment $\denv$ relates each
rule type in the earlier implicit environment to a System~F variable
$x$; %\bruno{This should be System F expression, right?};
% \bruno{In general it is an expression right? Not limited to variables.}\tom{
% The implicit abstraction rule always introduces a variable and there is no
% other rule that extends the implicit environment.
% }
this
variable serves as explicit value-level evidence for the implicit rule.
Lookup
in the translation environment is defined similarly to lookup in the
implicit environment, except that the lookup now returns a pair of a rule type
and an evidence variable.

The function $|\cdot|$ takes 
$\ourlang$ types $\rho$ to System F types T: 
\begin{center}
\colorbox{light}{\parbox{0.3\textwidth}{\begin{eqnarray*}
|\alpha| & = & \alpha \\
%%|\tyint| & = & \tyint \\
|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| \\
|\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
|\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| %\\
\end{eqnarray*}
% \end{mdframed}
}}
\end{center}

%Note that an empty context is translated to \tyunit{} in order to delay the
%evaluation of the rule.l
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
%\paragraph{Translating Empty Contexts}

% WC: I commented out the following paragraph because I thought this
% is too much
% For simplicity of presentation, Figure~\ref{f:trans} ignores the
% issue of empty contexts $\varnothing$. The simplified translation
% does not distinguish a rule with an empty context
% $\ruleabs{(\varnothing \To \type)}{e}$ from its body $e$.
% Nevertheless, in the operational semantics these two are different:
% a rule defers the evaluation of the rule body until explicit rule
% application. To properly capture this ``lazy'' evaluation strategy,
% we turn an empty context into an explicit \tyunit{} function
% parameter:

Variables, lambda abstractions and applications
are translated straightforwardly. Perhaps the only noteworthy 
rule is \TyIAbs. This rule associates the type $\rho_1$ with 
the fresh variable $x$ in the translation environment ($\denv$). 
This creates the necessary evidence that can be used by resolutions 
in the body of the rule abstraction to construct System F terms of type $|\rho_1|$. 
% The resolution process 
% will lookup the translation environment for a suitably matching rule. 
% If a suitable rule is found, the associated variable will then 
% be used to construct the evidence term.

\paragraph{Resolution}
The more interesting part of the translation happens when 
resolving queries. Queries are translated by rule $\TyQuery$ using the auxiliary
resolution judgment $\vturns$:
\begin{center}
$\bar{\alpha}; \env \vturns \rho~\gbox{\leadsto E}$
\end{center}

\noindent which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.
%%
%% defined by rule $\TrRes$\bruno{Need to add names 
%%to the figures!}. 
The translation of resolution basically extends the type-checking
process with simultaneously building System F evidence terms. The mechanism that builds
evidence dualizes the process of peeling off abstractions and universal 
quantifiers. The rule \mylabel{R-IAbs}~wraps a lambda binder with a fresh variable 
$x$ around a System F expression $E$, which is generated from the resolution 
for the head of the rule ($\rho_2$). The rule \mylabel{R-TAbs}~wraps a type lambda binder 
around the System F expression resulting from the resolution of $\rho$.
For simple types \mylabel{R-Simp}, evidence $E_1$ is retrieved from matching 
$\type$ and then used in the resolution process of the simple type 
$\type$ with rule $\turns_\downarrow$:

\begin{center}
$\bar{\alpha};\env; \rho~\gbox{\leadsto E_1} \turns_\downarrow \type~\gbox{\leadsto E_2}$
\end{center}

The rules \mylabel{I-IAbs}~and \mylabel{I-TAbs}~are the most interesting in the translation. 
In rule \mylabel{I-IAbs}~we need evidence for $\rho_2$ to build evidence for $\type$, 
but all we have is evidence $E_1$ for $\rho_1 \iarrow \rho_2$. However, we can 
construct evidence for $\rho_2$, by generating evidence $E_2$ for $\rho_1$
and then simply applying $E_1$ to $E_2$. Rule \mylabel{I-TAbs}~is similar: we need 
evidence for $\rho$, but all we have is evidence $E_1$ for $\forall \alpha. \rho$. 
To create evidence for $\rho$ we can substitute $\alpha$ by some type 
$\rho'$ in $\rho$ and generate a term which is the type application of 
$E_1$ to $|\rho'|$.

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
% Note how empty type contexts are also treated specially in the translation rules.
% They are translated respectively to $\tyunit$ abstractions and $\unit$ applications.

% \paragraph{Coherence of Translation}

% The type system of Section~\ref{s:types} imposes the condition
% $\welldefined(\env)$ to ensure coherence. Because translation performs
% resolution statically, it requires a stronger condition, $\stable{(\Delta)}$.
% Consider the following query:
% \begin{center}
%   $\myset{\relation{\forall \alpha. \alpha \to \alpha}{x}};
%   \myset{\relation{\tyint \to \tyint}{y}} \vturns \qask(\beta \to
%   \beta)$
% \end{center}
% where $\beta$ is a free type variable. By design, the operational semantics
% always chooses the most specific rule in the environment at runtime. For the
% sake of coherence, the \welldefined predicate in the type system
% guarantees that all possible runtime instantiations of $\beta\to\beta$ can be
% resolved.

% In contrast, the translation must decide statically on a single rule
% that is the most specific one for all possible instantiations.
% In the example, this decision cannot be made statically without knowing more about $\beta$. 
% The rule $\forall \alpha. \alpha\to\alpha$ covers all possible cases, but when
% $\beta$ is $\tyint$, it is not the most specific one. Hence, static resolution
% cannot be performed in a coherent manner. 
% The \stable{} predicate enforces static coherence, and is responsible for
% rejecting this problematic example.

% \paragraph{Incompleteness of the Translation}
% As the example above shows, not all well-typed $\ourlang$ programs
% have a translation; i.e. \\
% $\exists \denv. \welldefined(\dom(\denv)) \wedge \neg
% \stable(\denv)$. It is because $\ourlang$ does not respect
% parametricity, whereas System F does.

% From this remark on parametricity we can conclude that the operational
% semantics is a viable choice for systems where parametricity is not a
% concern and run-time type information is available. Mainstream
% object-oriented languages like Java and C\# fit this category.  On the
% other hand the elaboration semantics is a better match for functional
% programming languages concerned about parametricity (such as Haskell).

% As a remark we note that the elaboration semantics respects
% \emph{parametricity}~\cite{}, whereas the operational semantics does
% not. The example |f1| is a good representative of this fact. In
% System~F the only function that can inhabit the type |forall b . b ->
% b| is the identity function. As such, the selection of the most
% general instance in the two examples is the only possible choice for
% an elaboration into System~F. However, as we have just argued, this
% choice would be incoherent.

% The availability of run-time type information in the operational
% semantics, allows resolution to depend on that information. This 
% means that resolution can be more powerful and yet retain 
% coherence, but it also means that we lose parametricity. 
% Indeed, despite having the type |forall b . b -> b|, when applied 
% to an integer, the function |f1| is the successor (and not the
% identity) function.

%\paragraph{Correctness of Translation}

% The correctness of the translation is captured in two properties: type
% preservation and semantics preservation.
% Type preservation relates the type systems of $\ourlang$ and
% System~F. The translated expression of a well typed $\ourlang$
% expression is well typed.

\thmtranstypreserve
\begin{proof} (Sketch)
  We first prove\footnote{in the technical report} the more general lemma ``if $\tenv \mid \denv \turns
  \relation{e}{\rho} \leadsto E$, then $|\tenv|,|\denv| \turns
  \relation{E}{|\rho|}$'' by induction on the derivation of
  translation. Then, the theorem trivially follows.
\end{proof}

An important lemma in the theorem's proof is the type preservation of 
resolution.
\begin{lemma}[Type-Preserving Resolution]
Let $\env$ be an implicit environment, $\rho$ be a type and E be a System F expression.
If $\bar{\alpha};\env \vdash_r \rho \leadsto E$, then $|\env| \vdash E : |\rho|$.
\end{lemma}
% Both the theorem and the lemma are proven in Appendix~\ref{proof:preservation}.

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
\begin{lemma}[Determinacy]
The generated evidence of resolution is uniquely determined.
\[\forall \env, \rho, E_1, E_2, \bar{\alpha}: \quad\quad \bar{\alpha};\env \vdash_r \rho \leadsto E_1 ~\wedge~ \bar{\alpha};\env \vdash_r \rho \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
\end{lemma}
\begin{lemma}[Soundness]
Figure~\ref{fig:resolution2}'s definition of resolution (here denoted $\vdash_r^3$)
is sound (but incomplete) with respect to Figure~\ref{fig:resolution1}'s definition
(here denoted $\vdash_r^2$).
\[\forall \env, \rho, E: \quad\quad \env \vdash_r^3 \rho \leadsto E \quad\Rightarrow\quad \env \vdash_r^2 \rho \leadsto E \]
\end{lemma}
\begin{lemma}[Coherence]
Resolution is stable under substitution.
\[\forall \env, \rho, E, \bar{\alpha}, \theta: \quad\quad \bar{\alpha};\env \vdash_r \rho \leadsto E ~\wedge~ \mathit{dom}(\theta) \cap \bar{\alpha} = \emptyset \quad\Rightarrow\quad \bar{\alpha};\theta(\env) \vdash_r \theta(\rho) \leadsto \theta(E) \]
\end{lemma}

@@

The main difference, and complication, is due to the fact that the evidence for
type instantiation and recursive resolution is needed before these operations
actually take place, as the algorithm has to postpone them. For this reason, the
algorithm first produces placeholders that are later substituted for the actual
evidence.

The central relation is $\rho; \bar{\rho}; \bar{\alpha} \gbox{; \bar{\omega};
E}\turns_{\mathit{match}} \tau \hookrightarrow \bar{\rho}' \gbox{;
\bar{\omega}'; E'}$. It captures the matching instantiation of context type
$\rho$ against simple type $\tau$.  The input evidence for $\rho$ is $E$, and
the output evidence for the instantiation is $E'$. The accumulating parameters
$\bar{\alpha}$ and $\bar{\rho}$ denote that the instantiation of type variables
$\bar{\alpha}$ and the recursive resolution of $\bar{\rho}$ have been
postponed. We use the $\bar{\alpha}$ themselves as convenient placeholders for
the instantiating types, and we use the synthetic $\bar{\omega}$ as placeholders for
the evidence of the $\bar{\rho}$. The rules \mylabel{MTC-Abs} and \mylabel{MTC-Abs} 
introduce these two kinds of placeholders in the evidence. The former kind, $\bar{\alpha}$,
are substituted in rule \mylabel{MTC-Simp} where the actual type instantiatons $\bar{\theta}$
are computed. The latter kind, $\bar{\omega}$, are substituted later in rule \mylabel{Alg-Simp} where
the recursive resolutions take place.

Now we can state the correctness of the algorithm.
\begin{theorem}[Partial Correctness]
Let $\env$ be an implicit environment, $\rho$ be a type, E be a System F expression and $\bar{\alpha}$ a set of type variables.
Assume that $\epsilon \vdash_{\mathit{unamb}} \rho$ and also $\forall \rho_i \in \env: \epsilon \vdash_{\mathit{unamb}} \rho_i$.
Then $\bar{\alpha};\env \vdash_r \rho \leadsto E$ if and only if $\bar{\alpha};\env \vdash_{\mathit{alg}} \rho \leadsto E$,
provided that the algorithm terminates.
\end{theorem}

%-------------------------------------------------------------------------------
\subsection{Dynamic Semantics}
Finally, we define the dynamic semantics of $\ourlang$ as the composition
of the type-directed translation and System F's dynamic semantics. 
Following Siek's notation~\cite{systemfg}, this dynamic semantics is:
\[ \mathit{eval}(e) = V \quad\quad \textit{where } \epsilon\mid\epsilon \turns \relation{e}{\rho} \leadsto E \textit{ and } E \rightarrow^* V  \]
with $\rightarrow^*$ the reflexive, transitive closure of System F's standard single-step call-by-value reduction relation (see \cite[Chapter 23]{tapl}).

Now we can state the conventional type safety theorem for $\ourlang$:
\begin{theorem}[Type Safety]
If $\epsilon \mid \epsilon \turns \relation{e}{\rho}$, then $\mathit{eval}(e) = V$ for
some System F value $V$.
\end{theorem}
The proof follows trivially from Theorem~\ref{thm:type:preservation}.

>>>>

@@

%-------------------------------------------------------------------------------
\subsection{Type-Directed Translation}
Figure~\ref{fig:type} presents the translation rules that convert $\ourlang$
expressions into ones of System F. 
The gray parts of the figure extend the type system with the necessary
information for the translation.

@@

The gray extension to the syntax of type environments annotates very
implicit rule type with explicit System F evidence in the form of a 
term variable $x$.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Translation of Types}

The function $|\cdot|$ takes 
$\ourlang$ types $\rulet$ to System F types T: 
\begin{eqnarray*}
|\alpha| & = & \alpha \\
|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| \\
|\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
|\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| %\\
\end{eqnarray*}
Its reveals that implicit $\ourlang$ arrows are translated to explicit System F function arrows.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Translation of Terms}

The type-directed translation judgment, which extends the typing judgment, is 
\begin{center}
  $\tenv \turns \relation{e}{\rulet}~\gbox{\leadsto E}$
\end{center}
This judgment states that the translation of $\ourlang$ expression $e$ with
type $\rulet$ is System~F expression $E$, with respect to type environment
$\tenv$.

Variables, lambda abstractions and applications
are translated straightforwardly. Perhaps the only noteworthy 
rule is \TyIAbs. This rule associates the type $\rulet_1$ with 
the fresh variable $x$ in the type environment. 
This creates the necessary evidence that can be used by resolutions 
in the body of the rule abstraction to construct System F terms of type $|\rulet_1|$.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Resolution}
The more interesting part of the translation happens when resolving queries.
Queries are translated by rule $\TyQuery$ using the auxiliary resolution
judgment $\vturns$:
\begin{equation*}
\tenv \vturns \rulet~\gbox{\leadsto E}
\end{equation*}
which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.
The translation builds a System F term as evidence for the resolution.
The mechanism that builds evidence dualizes the process of peeling off
abstractions and universal 
quantifiers. The rule \mylabel{R-IAbs}~wraps a lambda binder with a fresh variable 
$x$ around a System F expression $E$, which is generated from the resolution 
for the head of the rule ($\rulet_2$). The rule \mylabel{R-TAbs}~wraps a type lambda binder 
around the System F expression resulting from the resolution of $\rulet$.

TODO

For simple types \mylabel{R-Simp}, evidence $E_1$ is retrieved from matching 
$\type$ and then used in the resolution process of the simple type 
$\type$ with rule $\turns_\downarrow$:

\begin{center}
$\bar{\alpha};\env; \rulet~\gbox{\leadsto E_1} \turns_\downarrow \type~\gbox{\leadsto E_2}$
\end{center}

The rules \mylabel{I-IAbs}~and \mylabel{I-TAbs}~are the most interesting in the translation. 
In rule \mylabel{I-IAbs}~we need evidence for $\rulet_2$ to build evidence for $\type$, 
but all we have is evidence $E_1$ for $\rulet_1 \iarrow \rulet_2$. However, we can 
construct evidence for $\rulet_2$, by generating evidence $E_2$ for $\rulet_1$
and then simply applying $E_1$ to $E_2$. Rule \mylabel{I-TAbs}~is similar: we need 
evidence for $\rulet$, but all we have is evidence $E_1$ for $\forall \alpha. \rulet$. 
To create evidence for $\rulet$ we can substitute $\alpha$ by some type 
$\rulet'$ in $\rulet$ and generate a term which is the type application of 
$E_1$ to $|\rulet'|$.

@@

\paragraph{Meta-Theory} The type-directed translation of $\ourlang$ to System F exhibits a number
of desirable properties.

\begin{theorem}[Type-preserving translation]\label{thm:type:preservation} Let $e$ be an $\ourlang$
  expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ be a System F expression. If
  $\tenv \turns \relation{e}{\rulet} \leadsto E$, then $|\tenv| \turns \relation{E}{|\rulet|}$.
\end{theorem}
Here we define the translation of the type environment form $\ourlang$ to System F as:
\begin{equation*}
\begin{array}{rcl@{\hspace{2cm}}rcl}
|\epsilon| & = & \epsilon & |\tenv,\alpha| & = & |\tenv|, \alpha \\
|\tenv,x : \rulet| & = & |\tenv|, x : |\rulet| &
|\tenv, \rulet \leadsto x| & = & |\tenv|, x : |\rulet|
\end{array}
\end{equation*}

An important lemma in the theorem's proof is the type preservation of 
resolution.
\begin{lemma}[Type-Preserving Resolution]
Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expression.
If $\tenv \vturns \rulet \leadsto E$, then $|\tenv| \vdash E : |\rulet|$.
\end{lemma}
% Both the theorem and the lemma are proven in Appendix~\ref{proof:preservation}.

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
\begin{lemma}[Determinacy]
The generated evidence of resolution is uniquely determined.
\[\forall \tenv, \rulet, E_1, E_2: \quad\quad \tenv \ivturns \rulet \leadsto E_1 ~\wedge~ \tenv \ivturns \rulet \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
\end{lemma}
\begin{lemma}[Soundness]
Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
\[\forall \tenv, \rulet, E: \quad\quad \tenv \ivturns \rulet \leadsto E \quad\Rightarrow\quad \tenv \vturns \rulet \leadsto E \]
\end{lemma}
\begin{lemma}[Coherence]
Resolution is stable under substitution.
\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \quad\quad 
\tenv,\alpha,\tenv' \ivturns \rulet \leadsto E \quad\wedge\quad \tenv \vdash \sigma
\quad\Rightarrow\quad 
\tenv,\tenv'[\sigma/\alpha] \ivturns \rulet[\sigma/\alpha] \leadsto E[|\sigma|/\alpha] \]
\end{lemma}

@@

% The main difference, and complication, is due to the fact that the evidence for
% type instantiation and recursive resolution is needed before these operations
% actually take place, as the algorithm has to postpone them. For this reason, the
% algorithm first produces placeholders that are later substituted for the actual
% evidence.
% 
% The central relation is $\rulet; \bar{\rulet}; \bar{\alpha} \gbox{; \bar{\omega};
% E}\turns_{\mathit{match}} \tau \hookrightarrow \bar{\rulet}' \gbox{;
% \bar{\omega}'; E'}$. It captures the matching instantiation of context type
% $\rulet$ against simple type $\tau$.  The input evidence for $\rulet$ is $E$, and
% the output evidence for the instantiation is $E'$. The accumulating parameters
% $\bar{\alpha}$ and $\bar{\rulet}$ denote that the instantiation of type variables
% $\bar{\alpha}$ and the recursive resolution of $\bar{\rulet}$ have been
% postponed. We use the $\bar{\alpha}$ themselves as convenient placeholders for
% the instantiating types, and we use the synthetic $\bar{\omega}$ as placeholders for
% the evidence of the $\bar{\rulet}$. The rules \mylabel{MTC-Abs} and \mylabel{MTC-Abs} 
% introduce these two kinds of placeholders in the evidence. The former kind, $\bar{\alpha}$,
% are substituted in rule \mylabel{MTC-Simp} where the actual type instantiatons $\bar{\theta}$
% are computed. The latter kind, $\bar{\omega}$, are substituted later in rule \mylabel{Alg-Simp} where
% the recursive resolutions take place.

With the evidence we can state the correctness of the algorithm.
\begin{theorem}[Partial Correctness]
Let $\tenv$ be a type environment, $\rulet$ be a type and E be a System F expressions.
Assume that $\epsilon \vdash_{\mathit{unamb}} \rulet$ and also $\forall \rulet_i \in \tenv: \epsilon \vdash_{\mathit{unamb}} \rulet_i$.
Then $\tenv \ivturns \rulet \leadsto E$ if and only if $\tenv \vdash_{\mathit{alg}} \rulet \leadsto E$,
provided that the algorithm terminates.
\end{theorem}

%-------------------------------------------------------------------------------
\subsection{Dynamic Semantics}
Finally, we define the dynamic semantics of $\ourlang$ as the composition of
the type-directed translation and System F's dynamic semantics.  Following
Siek's notation~\cite{systemfg}, this dynamic semantics is:
\[ \mathit{eval}(e) = V \quad\quad \textit{where } \epsilon \turns \relation{e}{\rulet} \leadsto E \textit{ and } E \rightarrow^* V  \]
with $\rightarrow^*$ the reflexive, transitive closure of System F's standard single-step call-by-value reduction relation (see \cite[Chapter 23]{tapl}).

Now we can state the conventional type safety theorem for $\ourlang$:
\begin{theorem}[Type Safety]
If $\epsilon \turns \relation{e}{\rulet}$, then $\mathit{eval}(e) = V$ for
some System F value $V$.
\end{theorem}
The proof follows trivially from Theorem~\ref{thm:type:preservation}.
